Anxiety, avoidance, and sequential evaluation

### 2019-11-17 manuscripts/manuscript/manuscript.tex

Though anxiety disorders differ in their particular symptomology, and in the content and situations that elicit symptoms, they all are similarly characterized by aberrations in the processing of and response to threat \citep{dsm5}. In particular, at least three symptoms manifest across the anxiety disorders. First, anxiety is associated with exaggerated threat appraisal, or a bias towards evaluating threat as disproportionately greater in likelihood and severity than is warranted \citep{ClarkBeck2011}. Second, anxiety is also associated with fear generalization, wherein the primary threat becomes associated with increasingly distal locations, events, and thoughts \citep{dymond2015}. Finally, anxiety is associated with persistent avoidance behavior, which often occurs well in advance of the materialization of actual threat \citep{Arnaudova2017}. Excessive avoidance behaviors are an especially harmful aspect of anxiety disorders, both because they interfere with daily life and because they indirectly maintain anxiety by preventing learning from the non-occurrence of perceived threats. Though laboratory studies of decision making and learning in anxious populations have corroborated these clinical observations \citep{Harle2017, norbury2018, Aylward2019}, none offer an explanation as to the root of these symptoms.

@@

This line of reasoning hints that a fundamental deficit in anxiety disorders may relate to this assumption, which otherwise should preclude the spread of threat to antecedent situations and subsequent excessive avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future \citep{ClarkBeck2011}. Clinically and subclinically anxious individuals judge future threat as more likely than do non-anxious individuals\citep{ButlerMathews1983, ButlerMathews1987, MacleodByrne1996}. Importantly, the development and maintenance of clinical anxiety is strongly tied to diminished perceived control\citep{bandura1977, barlow2002, gallagher2014a}, such that anxious individuals are more likely to endorse the belief that they are unlikely or unable to mitigate future threat. Indeed, lack of belief in one's ability to successfully navigate future danger is associated with anxiety \citep{davey1996, dugas1997}, and an increased belief in perceived control over threat is correlated with symptom reduction across the family of anxiety disorders\citep{gallagher2014b}.

@@

\begin{equation*}
Q^\pi(s,a) = r(s,a) + \gamma \sum_{s'} p(s' \mid s,a) \sum_{a'} \pi(a' \mid s') Q^\pi(s',a').
\end{equation*}

@@

Relatedly, the model can also capture findings of increased behavioral inhibition, measured as prolonged response times, in anxious individuals under threat\cite{bach2015}. In the behavioral inhibition task, participants seek tokens adjacent to a virtual "sleeping predator," which are all lost if the predator awakes. Though in this variant the risk of predation is constant throughout a trial (rather than increasing, as in the BART) the potential loss from capture still increases with each token collected. Bach \cite{bach2015} finds that participants are slower to collect tokens as this potential loss increases, and that this slowing is enhanced by subclinical anxiety. We can capture this effect in our model by noting that the relative value of approach compared to avoidance is reduced as potential loss and the risk that the predator awakes increase (Fig.~2d/e). These effects are amplified under more pessimistic (i.e., in our model, more anxious) assumptions about future actions. Thus, as before, anxiety in our model produces greater and earlier choice of avoidance, analogous to earlier cash-out in the BART. To extend this effect to reaction times, we adopt the further, standard assumption that actions (here, approach) are slower when their relative value compared to alternatives is lower \citep{oud2016}. In this case, the model captures behavioral inhibition (slower responses as threat increases), and its enhancement by anxiety, as measured by Bach. (Note that we need not assume the coupling of reaction times to action value spread is due to difficulty in decision formation per se, which Bach argues against: it may, for instance, reflect Pavlovian initiation biases \cite{niv2007}.)

@@

Our model predicts this result (Fig.~3b), though for a somewhat different reason. In our model, pessimistic (anxious) agents neglect large gains deeper in the tree, not because they fail to consider them (here we assume full evaluation of the Bellman equation), but because with increasing anxiety they increasingly expect the potential of choosing incorrectly afterward, thus failing to recoup the loss (and, mathematically, probabilistically pruning the better branches). Future research could use slight variants in the decision trees to tease apart these different interpretations.

@@

@@

We have centered our discussion at Marr's \cite{marr1982} computational level: on beliefs and their consequences in terms of action values. We have so far remained agnostic as to how, algorithmically or mechanistically, these misbeliefs are implemented in the brain. \cite{friston2014computational} Importantly, the brain is believed to contain multiple distinct mechanisms for evaluating actions (e.g, model-based and model-free learning; \cite{daw2005, huys2015}), and pessimisitic beliefs might play out either differentially or similarly through each of these mechanisms. One promising possibility is that these symptoms mainly reflect aberrations in model-based planning \cite{huys2015}, i.e. explicitly evaluating actions by mentally simulating possible trajectories. Recent work suggests this process may be accomplished by mentally ``replaying'' individual potential trajectories \cite{momennejad_offline_2018,mattar2018}. In this setting, the biases we suggest would amount to over-contemplating, or over-weighting, certain pessimistic trajectories \cite{hunter2019}. Such a bias might be detectable using neuroimaging, as a change in which types of events that tend to be replayed \cite{momennejad_offline_2018, ambrose2016}. Such a biased replay process, in turn, may also correspond to worry and rumination. Indeed, in line with the present results, chronic worry is associated with reduced perceived control, diminished belief in self-efficacy in response to threat, and exaggerated threat appraisal\cite{Berenbaum2010}. This suggests that clinical anxiety may in part result from planning processes gone awry.

It is important to note that the present model may not describe all anxiety disorders with equal accuracy. Indeed, our analysis of pessimistic sequential evaluation is, by definition, a model of prospective cognition. Thus, the present results are more likely to accurately describe the anxiety disorders which primarily involve aberrations in future-oriented cognitive processes, such as generalized and social anxiety disorders. Naturally, the present model can account neither for the compulsive behaviors of obsessive-compulsive disorder (OCD) nor the memory disturbances of post-traumatic stress disorder (PTSD). That said, recent clinical studies suggest that diminished perceived control is a vulnerability factor common to all anxiety disorders, including OCD and PTSD \citep{gallagher2014a, gallagher2014b}. Importantly, aspects of other psychiatric disorders that involve future-oriented misbeliefs, worry, and avoidance behaviors (e.g. eating disorders\citep{konstantellou2011}) may similarly be well-described by the current account. As such, the present results are transdiagnostic and not limited to one particular diagnosis. Finally, the relationship between evaluation, planning, and neural replay discussed above suggests potential future work that might help to bring this model into contact with the more memory-related aspects of PTSD. For instance, the same replay processes that can be used to evaluate actions can also, in theory, update predictive representations, cognitive maps, or models of the environment \citep{russek_predictive_2017}, such as the successor representation \cite{dayan1993, momennejad_successor_2017}. If so, similar biases in replay could result in not just aberrant avoidance behavior, but also progressive, aberrant remodeling of world models or cognitive maps, an observation which may connect to the rich and complex set of issues on memory involvement in PTSD.

>>>>

Though anxiety disorders differ in their particular symptomology, and in the content and situations that elicit symptoms, they all are similarly characterized by aberrations in the processing of and response to threat \citep{dsm5}. In particular, at least three symptoms manifest across many of the anxiety disorders. First, anxiety is associated with exaggerated threat appraisal, or a bias towards evaluating threat as disproportionately greater in likelihood and severity than is warranted \citep{ClarkBeck2011}. Second, anxiety is also associated with fear generalization, wherein the primary threat becomes associated with increasingly distal locations, events, and thoughts \citep{dymond2015}. Finally, anxiety is associated with persistent avoidance behavior, which often occurs well in advance of the materialization of actual threat \citep{Arnaudova2017}. Excessive avoidance behaviors are an especially harmful aspect of anxiety disorders, both because they interfere with daily life and because they indirectly maintain anxiety by preventing learning from the non-occurrence of perceived threats. Though laboratory studies of decision making and learning in anxious populations have corroborated these clinical observations \citep{Harle2017, norbury2018, Aylward2019}, none offer an explanation as to the root of these symptoms.

@@

This line of reasoning hints that a fundamental aberration in anxiety disorders may relate to this assumption, which otherwise should preclude the spread of threat to antecedent situations and subsequent excessive avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future \citep{ClarkBeck2011}. Clinically and subclinically anxious individuals judge future threat as more likely than do non-anxious individuals\citep{ButlerMathews1983, ButlerMathews1987, MacleodByrne1996}. Importantly, the development and maintenance of clinical anxiety is strongly tied to diminished perceived control\citep{bandura1977, barlow2002, gallagher2014a}, such that anxious individuals are more likely to endorse the belief that they are unlikely or unable to mitigate future threat. Indeed, lack of belief in one's ability to successfully navigate future danger is associated with anxiety \citep{davey1996, dugas1997}, and an increased belief in perceived control over threat is correlated with symptom reduction across the family of anxiety disorders\citep{gallagher2014b}.

@@

\begin{equation}\label{eq:bellman_full}
Q^\pi(s,a) = r(s,a) + \gamma \sum_{s'} p(s' \mid s,a) \sum_{a'} \pi(a' \mid s') Q^\pi(s',a').
\end{equation}

@@

Relatedly, the model can also capture findings of increased behavioral inhibition, measured as prolonged response times, in anxious individuals under threat\cite{bach2015}. In the behavioral inhibition task, participants seek tokens adjacent to a virtual "sleeping predator", which are all lost if the predator awakes. Though in this variant the risk of predation is constant throughout a trial (rather than increasing, as in the BART) the potential loss from capture still increases with each token collected. Bach \cite{bach2015} finds that participants are slower to collect tokens as this potential loss increases, and that this slowing is enhanced by subclinical anxiety. We can capture this effect in our model by noting that the relative value of approach compared to avoidance is reduced as potential loss and the risk that the predator awakes increase (Fig.~2d/e). These effects are amplified under more pessimistic (i.e., in our model, more anxious) assumptions about future actions. Thus, as before, anxiety in our model produces greater and earlier choice of avoidance, analogous to earlier cash-out in the BART. To extend this effect to reaction times, we adopt the further, standard assumption that actions (here, approach) are slower when their relative value compared to alternatives is lower \citep{oud2016}. In this case, the model captures behavioral inhibition (slower responses as threat increases), and its enhancement by anxiety, as measured by Bach. (Note that we need not assume the coupling of reaction times to action value spread is due to difficulty in decision formation per se, which Bach argues against: it may, for instance, reflect Pavlovian initiation biases \cite{niv2007}.)

@@

Our model predicts this result (Fig.~3b), though for a somewhat different reason. In our model, pessimistic (anxious) agents neglect large gains deeper in the tree, not because they fail to consider them (here we assume full evaluation of the Bellman equation), but because with increasing anxiety they increasingly expect the potential of choosing incorrectly afterward, thus failing to recoup the loss (and, mathematically, probabilistically pruning the better branches). Future research could use slight variants in the decision trees to tease apart these different interpretations. Tree order matters in delineating between these theories. What's one good example?

@@

Insert paragraph here about testing. Delgado experiment for one. Learning transitions another?

@@

We have centered our discussion at Marr's \cite{marr1982} computational level: on beliefs and their consequences in terms of action values. We have so far remained agnostic as to how, algorithmically or mechanistically, these misbeliefs are implemented in the brain. \cite{friston2014computational} Importantly, the brain is believed to contain multiple distinct mechanisms for evaluating actions (e.g, model-based and model-free learning; \cite{daw2005, huys2015}), and pessimistic beliefs might play out either differentially or similarly through each of these mechanisms. One promising possibility is that these symptoms mainly reflect aberrations in model-based planning \cite{huys2015}, i.e. explicitly evaluating actions by mentally simulating possible trajectories. Recent work suggests this process may be accomplished by mentally ``replaying'' individual potential trajectories \cite{momennejad_offline_2018,mattar2018}. In this setting, the biases we suggest would amount to over-contemplating, or over-weighting, certain pessimistic trajectories \cite{hunter2019}. Such a bias might be detectable using neuroimaging, as a change in which types of events that tend to be replayed \cite{momennejad_offline_2018, ambrose2016}. Such a biased replay process, in turn, may also correspond to worry and rumination. Indeed, in line with the present results, chronic worry is associated with reduced perceived control, diminished belief in self-efficacy in response to threat, and exaggerated threat appraisal\cite{Berenbaum2010}. This suggests that clinical anxiety may in part result from planning processes gone awry.

It is important to note that the present model may not describe all anxiety disorders with equal accuracy. Indeed, our analysis of pessimistic sequential evaluation is, by definition, a model of prospective cognition. Thus, the present results are more likely to accurately describe the anxiety disorders which primarily involve aberrations in future-oriented cognitive processes, such as generalized and social anxiety disorders. Naturally, the present model can account neither for the compulsive behaviors of obsessive-compulsive disorder (OCD) nor the memory disturbances of post-traumatic stress disorder (PTSD). That said, recent clinical studies suggest that diminished perceived control is a vulnerability factor common to all anxiety disorders, including OCD and PTSD \citep{gallagher2014a, gallagher2014b}. Importantly, aspects of other psychiatric disorders that involve future-oriented misbeliefs, worry, and avoidance behaviors (e.g. eating disorders\citep{konstantellou2011}) may similarly be well-described by the current account. Speculatively, the model may also have implications for understanding dysregulated goal pursuit in mania and bipolar disorder. Manic symptoms are associated with increases in perceived control \citep{
implications for understanding increased perceived control in mania \cite{}. As such, the present results are transdiagnostic and not limited to one particular diagnosis. Finally, the relationship between evaluation, planning, and neural replay discussed above suggests potential future work that might help to bring this model into contact with the more memory-related aspects of PTSD. For instance, the same replay processes that can be used to evaluate actions can also, in theory, update predictive representations, cognitive maps, or models of the environment \citep{russek_predictive_2017}, such as the successor representation \cite{dayan1993, momennejad_successor_2017}. If so, similar biases in replay could result in not just aberrant avoidance behavior, but also progressive, aberrant remodeling of world models or cognitive maps, an observation which may connect to the rich and complex set of issues on memory involvement in PTSD.
